State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-10-28 StateOfPractice/Papers/SOP_MI_CritReviewsBioMedEng/SOP_MI_OSR.tex
Initial draft of Critical Reviews in BioMed Eng paper

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with Harvard style bibliographic references

%\documentclass[preprint,12pt,authoryear]{elsarticle}
%\documentclass[3p, 12pt,authoryear]{elsarticle}
\documentclass[final, 3p, times, authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath,amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

@@

%\usepackage{tikz}
%\usetikzlibrary{mindmap}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{arydshln}
\usepackage{paralist}
\usepackage{booktabs}
\usepackage{pbox}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\mn}[1]{\authornote{magenta}{MN}{#1}} %Mike Noseworthy
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

\newcommand{\notdone}[1]{\textcolor{red}{#1}}
\newcommand{\done}[1]{\textcolor{black}{#1}}

@@

\journal{Academic Radiology}

\pgfplotsset{compat=1.13}

@@

\title{State of the Practice for Medical Imaging Software}

@@

\author[CAS]{Spencer Smith}
\author[CAS]{Ao Dong}
\author[CAS]{Jacques Carette}
\author[ECE]{Mike Noseworthy}

\affiliation[CAS]{organization={McMaster University, Computing and Software Department},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ECE]{organization={McMaster University, Electrical Engineering},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

@@

We present the state of the practice for Medical Imaging (MI) software. We
selected 29 medical imaging projects from 48 candidates, assessed 10 software
qualities (installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability,
visibility/transparency and reproducibility) by answering 108 questions for each
software project, and interviewed 8 of the 29 development teams. Based on the
quantitative data for the first 9 qualities, we ranked the MI software with the
Analytic Hierarchy Process (AHP). The four top ranked software products are:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
Our ranking is mostly consistent with the community's ranking, with four of our
top five projects also appearing in the top five of a list ordered by
stars-per-year. Generally, MI software is in a healthy state as shown by the
following: in the repositories we observed 88\% of the documentation artifacts
recommended by research software development guidelines, 100\% of MI projects
use version control tools, and developers appear to use the common quasi-agile
research software development process. However, the current state of the
practice deviates from the existing guidelines because of the rarity of some
recommended artifacts (like test plans, requirements specification, code of
conduct, code style guidelines, product roadmaps, and Application Program
Interface (API) documentation), low usage of continuous integration (17\% of the
projects), low use of unit testing (about 50\% of projects), and room for
improvement with documentation (six of nine developers felt their documentation
wasn't clear enough). From interviewing the developers, we identified five pain
points and two qualities of potential concern: lack of development time, lack of
funding, technology hurdles, ensuring correctness, usability, maintainability,
and reproducibility. The interviewees proposed strategies to improve the state
of the practice, to address the identified pain points, and to improve software
quality. Combining their ideas with ours, we have the following list of
recommendations: increase documentation, increase testing by enriching datasets,
increase continuous integration usage, move to web applications, employ linters,
use peer reviews, design for change, add assurance cases, and incorporate a
``Generate All Things'' approach.

@@

\begin{keyword}
	medical imaging, research software, software engineering, software
	quality, Analytic Hierarchy Process, developer interviews
\end{keyword}

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software.  MI tools use images of the interior of the body (from sources
such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
Zhang2008}.  Figure~\ref{Fig_Example}, which shows an image of the brain,
highlights the importance and value of MI. Through MI medical practitioners and
researchers can noninvasively gain insights into the human body, including
information on injuries and illnesses. Given the importance of MI software and
the high number of competing software projects, we wish to understand the merits
and drawbacks of the current development processes, tools, and methodologies. We
aim to assess through a software engineering lens the quality of the existing
software with the hope of highlighting standout examples, understanding current
pain points and providing guidelines and recommendations for future development.

\begin{figure}[!ht]
    \begin{center}
        \includegraphics[scale=0.25]{figures/MPR.png}        
    \end{center}
    \caption{Example brain image showing a multi-planar reformat using Horos
	(free open-source medical imaging/DICOM viewer for OSX, based on OsiriX)}
    \label{Fig_Example}
\end{figure}

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the development of research software in
general. We wish to understand the impact of the often cited gap, or chasm,
between software engineering and research software \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

To gain insights, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \citep{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, software artifacts, tools,
principles, processes, methodologies, and qualities.  When we mention artifacts
we mean the documents, scripts and code that constitutes a software development
project. Example artifacts include requirements, specifications, user manuals,
unit tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.  Our aim is to collect
enough information to understand the current pain points experienced by the MI
software development community so that we can make some preliminary
recommendations for future improvements.

We based the structure of the paper on the research questions, so for each
research question below we point to the section that contains our answer.  We
start with identifying the relevant examples of MI software for the assessment
exercise:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
	software projects exist, with the constraint that the source code must be
	available for all identified projects? (Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	(Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
	(Section~\ref{Sec_VsCommunityRanking})
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do MI projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do MI projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do MI projects compare to research software in general with respect to
	principles, processes, and methodologies used?
	(Section~\ref{Sec_CompareMethodologies})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on MI software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from MI compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
	developers what specific best practices are taken to address the pain points
	and software quality concerns? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practice could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}?
	(Section~\ref{ch_recommendations})

@@

To make the project feasible, we only cover MI visualization software.  As a
consequence we are excluding many other categories of MI software, including
Segmentation, Registration, Visualization, Enhancement, Quantification,
Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
and Communication) (as summarized by \citet{Bankman2000} and
\citet{Angenent2006}).  We also exclude Statistical Analysis and Image-based
Physiological Modelling \citep{enwiki:1034877594} and Feature Extraction,
Classification, and Interpretation \citep{Kim2011}. Software that provides MI
support functions is also out of scope; therefore, we have not assessed the
toolkit libraries VTK \citep{SchroederEtAl2006} and ITK \citep{McCormick2014}.
Finally, Picture Archiving and Communication System (PACS), which helps users to
economically store and conveniently access images \citep{Choplin1992}, are
considered out of scope.

\subsection{Methodology Overview}

We designed a general methodology to assess the state of the practice for
research software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be
found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{SmithEtAl2022,
Michalski2021}.  This methodology builds off prior work to assess the state of
the practice for such domains as Geographic Information Systems
\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
software \citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. We consider a person month as $20$ working days
($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose a research software domain (in the current
case MI) and identify a list of about 30 software packages. (For measuring MI we
used 29 software packages.)  We approximately measure the qualities of each
package by filling in a grading template. Compared with our previous
methodology, the new methodology also includes repository based metrics, such as
the number of files, number of lines of code, percentage of issues that are
closed, etc.  With the quantitative data in the grading template, we rank the
software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
provides details). After this, as another addition to our previous methodology,
we interview some development teams to further understand the status of their
development process.

\section{Background} \label{ch_background}

To measure the existing MI software we need two sets of definitions: i) the
definitions of relevant software license models (Section
\ref{sec_software_categories}); and, ii) the definitions of the software
qualities that we will be assessing (Section \ref{sec_software_quality}). In our
assessment we rank the software packages for each quality; therefore, this
section also provides the background on our ranking process --- the Analytic
Hierarchy Process (Section \ref{sec_AHP}).

\subsection{Software Categories} \label{sec_software_categories}

When assessing software packages, we need to know the software's license.  In
particular, we need to know whether the source code will be available to us or
not.  We define three common software categories.  We will only assess software
that fits under the Open Source Software license.

\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
accessible. Users have the right to study, change and distribute it under a
license granted by the copyright holder. For many OSS projects, the development
process relies on the collaboration of different contributors worldwide
\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
software project, such as the underlying logic of software functions, how
developers achieve their works, and the flaws and potential risks in the final
product. Thus, OSS is suitable for researchers analyzing the qualities of a
project.

\item \textbf{Freeware} Freeware is software that can be used free of charge.
Unlike OSS, the authors of do not allow access or modify the source code
\citep{LINFO2006}. To many end-users, the differences between freeware and OSS
may not be relevant. However, software developers who wish to modify the source
code, and researchers looking for insight into software development process may
find the inaccessible source code a problem.

\item \textbf{Commercial Software} ``Commercial software is software developed
by a business as part of its business'' \citep{GNU2019}. Typically speaking,
commercial software requires users to pay to access all of its features,
excluding access to the source code. However, some commercial software is also
free of charge \citep{GNU2019}. Based on our experience, most commercial
software products are not OSS.

\subsection{Software Quality Definitions} \label{sec_software_quality}

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  Below we list the 10 qualities of interest for this study.
The order of the qualities follows the order used in \citet{GhezziEtAl2003},
which puts related qualities (like correctness and reliability) together.
Moreover, the order is roughly the same as the order developers consider
qualities in practice.

\begin{itemize}
	\item \textbf{Installability} The effort required for the installation
    and/or uninstallation of software in a specified environment
    \citep{ISO/IEC25010, lenhard2013measuring}.

	\item \textbf{Correctness \& Verifiability} A program is correct if it
    matches its specification \citep[p.\ 17]{GhezziEtAl2003}.  The specification
    can either be explicitly or implicitly stated.  The related quality of
    verifiability is the ease with which the software components or the
    integrated product can be checked to demonstrate its correctness.

	\item \textbf{Reliability} The probability of failure-free operation of a
	computer program in a specified environment for a specified time \citep{musa1987software}, \citep[p.\ 357]{GhezziEtAl2003}.

	\item \textbf{Robustness} Software possesses the characteristic of
	robustness if it behaves ``reasonably'' in two situations: i) when it
	encounters circumstances not anticipated in the requirements specification,
	and ii) when users violate the assumptions in its requirements specification 
	\citep[p.\ 19]{GhezziEtAl2003}, \citep{boehm2007software}.

	\item \textbf{Usability} ``The extent to which a product can be used by
	specified users to achieve specified goals with effectiveness, efficiency,
	and satisfaction in a specified context of use'' \citep{ISO/TR16982:2002}
	\citep{ISO9241-11:2018}.

	\item \textbf{Maintainability} The effort with which a software system or
	component can be modified to i) correct faults; ii) improve performance or
	other attributes; iii) satisfy new requirements
	\citep{IEEEStdGlossarySET1990}, \citep{boehm2007software}.

	\item \textbf{Reusability} ``The extent to which a software component can be
	used with or without adaptation in a problem solution other than the one for
	which it was originally developed'' \citep{kalagiakos2003non}.

	\item \textbf{Understandability} ``The capability of the software product to
	enable the user to understand whether the software is suitable, and how it
	can be used for particular tasks and conditions of use'' \citep{iso2001iec}.

	\item \textbf{Visibility/Transparency} The extent to which all the steps
	of a software development process and the current status of it are conveyed
	clearly \citep[p.\ 32]{GhezziEtAl2003}.

	\item \textbf{Reproducibility} ``A result is said to be reproducible if
	another researcher can take the original code and input data, execute it,
	and re-obtain the same result'' \citep{BenureauAndRougier2017}.
\end{itemize}

\subsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}

>>>>

\documentclass[final, 12pt, 3p, times]{elsarticle}

\usepackage{amssymb}
\usepackage{amsmath}

@@

\usepackage{multirow}
\usepackage{booktabs}
\usepackage{pbox}
\usepackage{paralist}

@@

% prepared for: Journal of Imaging Informatics in Medicine
% https://link.springer.com/journal/10278

\journal{Critical Reviews in Biomedical Engineering}

@@

\title{State of the Practice for Medical Imaging Software Based on Open Source
Repositories}

@@

%\author{Anonymous}
\author[CAS]{Spencer Smith}
\author[CAS]{Ao Dong}
\author[CAS]{Jacques Carette}
\author[EE]{Michael Noseworthy}

\affiliation[CAS]{organization={McMaster University, Computing and Software
Department}, %Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ME]{organization={McMaster University, Electrical and Computer Engineering}, %Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

@@

    We present the state of the practice for Medical Imaging (MI) software based
on data available in open source repositories. We selected 29 projects from 48
candidates and assessed 9 software qualities (installability, correctness/
verifiability, reliability, robustness, usability, maintainability, reusability,
understandability, and visibility/transparency) by answering 108 questions for
each. Using the Analytic Hierarchy Process (AHP) on the quantitative data, we
ranked the MI software.  The top five are \textit{3D Slicer}, \textit{ImageJ},
\textit{Fiji}, \textit{OHIF Viewer}, and \textit{ParaView}.  This is consistent
with the community's view, with four of these also appearing in the top five
using GitHub metrics (stars-per-year).  
In general, the quality and quantity of documentation present in a project
correlates quite well with its popularity.

@@

\begin{keyword}
	medical imaging, research software, software engineering, software
	quality, analytic hierarchy process
\end{keyword}

@@

We study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide critical information for diagnostic, analytic, and medical
applications. Given its importance, we want to understand the merits and
drawbacks of the current development processes, tools, and methodologies. We
use a software engineering lens to assess the quality of existing MI software.

@@

As well as state of the practice (SOP) for MI software, we would like to
understand the impact of the often cited gap (chasm!) between recommended
software engineering practices and that used for most research 
software~\cite{Storer2017}. Although scientists spend
a substantial proportion of their working times on software development
\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
\textbf{open source} software projects exist? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\end{enumerate}

We only cover MI visualization software.  We exclude other categories of MI
software, including Segmentation, Registration, Visualization, Enhancement,
Quantification, Simulation, plus MI archiving and telemedicine systems
(Compression, Storage, and Communication).  We also exclude Statistical Analysis
and Image-based Physiological Modelling and Feature Extraction, Classification,
and Interpretation. Software that provides MI support functions is also out of
scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
Finally, Picture Archiving and Communication System (PACS), which helps users to
economically store and conveniently access images, are considered out of scope.

\subsection{Methodology} \label{SecMethodology}

We have a standard set of questions designed to assess the qualities of any
research software project.  [blind review redacted details on the author's
previous studies using the same methodology.]
%% DOUBLE BLIND
% ~\cite{SmithEtAl2021, SmithAndMichalski2022}.  This has
% been applied to MI software and Lattice Boltzmann Solvers \cite{SmithEtAl2024}.
% This builds off prior work to assess the state of the practice for such domains
% as Geographic Information Systems \cite{smith2018state}, Mesh Generators
% \cite{smith2016state}, Seismology software \cite{Smith2018Seismology}, and
% Statistical software for psychology \cite{smith2018statistical}.  
We maintain
the previous constraint that the work load for measuring a given domain should
take around one person-month's worth of effort ($20$ working days at $8$
person-hours per day).

We identify a list of potential packages (through online searches) which is then
filtered and vetted by a domain expert. We aim for roughly $30$ packages. For
each remaining package, we measure its qualities by filling in a grading
template [citation redacted for double blind]. %%DOUBLE BLIND
%~\cite{SmithEtAl2021}.  
This data is used to rank the projects with the
Analytic Hierarchy Process (AHP).  We summarize further details on the
interaction with the domain expert, software qualities, grading the software and
AHP below and in longer form in [redacted for double blind].
%% DOUBLE BLIND
%Smith et al.\ (2024)~\cite{SmithEtAl2024_MI_SOP}.

\subsubsection{Domain Expert} \label{sec_vet_software_list}

The Domain Expert vets the proposed list because online resources can be
inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
our Domain Expert is [details of our domain expert removed for double-blind].
%%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
%University, Co-Director of the McMaster School of Biomedical Engineering, and
%Director of Medical Imaging Physics and Engineering at St.\ Joseph's
%Healthcare, Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, they are asked to
independently create a list of top software packages in the domain.  This helps
get the expert's knowledge refreshed in advance of the meeting.

\subsubsection{Software Qualities} \label{sec_software_quality}

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as
a set of measures.  That is, quality is a collection of different qualities,
often called ``ilities.''  For this study we selected 9 qualities to measure:
installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, and visibility/transparency.
With the exception of installability, all the qualities are defined in Ghezzi
et al. (2003) \cite{GhezziEtAl2003}. Installability is defined as the effort
required for the installation and/or uninstallation of software in a specified
environment \cite{ISO/IEC25010}.

\subsubsection{Grading} \label{sec_grading_software}

We use an existing template [citation redacted]
%% DOUBLE BLIND
%~\cite{SmithEtAl2021} 
that is designed to measure the
aforementioned qualities. To stay within our given measurement time frame, each
package gets up to five hours of time.  Project developers can be contacted for
help regarding installation, if necessary, but we impose a cap of about two
hours on the installation process.  Figure~\ref{fg_grading_template_example}
shows an excerpt of the measurement spreadsheet.  The rows are the measures and
the columns correspond to the software packages.  [The full data is available on
Mendeley; link will be provided after refereeing.] 
% \cite{Dong2021-Data}
%provides the full set of measurement data.   %%DOUBLE BLIND

\begin{figure*}[!ht]
\includegraphics[scale=0.53]{template.pdf}
\caption{Grading template example}
\label{fg_grading_template_example}
\end{figure*}

The full template consists of 108 questions over 9 qualities.  These
questions are designed to be unambiguous, quantifiable, and measurable
with constrained time and domain knowledge.

The grader, after answering questions for each quality assigns an overall
score (between 1 and 10) based on the answers.  Several of the qualities use
the word ``surface'' to highlight that these particular qualities are a shallow
measure.  For example, usability is not measured using user studies.
Instead, we look for signs that the developers considered usability.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
details on quality measurement are provided in [redacted for double blind].
%% DOUBLE BLIND
%Smith et al.\ (2024) \cite{SmithEtAl2024_MI_SOP}.

\subsubsection{Analytic Hierarchy Process (AHP)} \label{sec_AHP}

Developed by Saaty in the 1970s, AHP is widely used to analyze multiple criteria
decisions~\cite{VaidyaEtAl2006}. AHP organizes multiple criteria in a
hierarchical structure and uses pairwise comparisons between alternatives to
calculate relative ratios~\cite{Saaty1990}. AHP works with sets of $n$
\textit{options} and $m$ \textit{criteria}.  In our project $n=29$ and $m=9$
since there are 29 options (software products) and 9 criteria (qualities). With
AHP the sum of the grades (scores) for all products for a given quality will be
1.0.  We rank the software for each of the qualities, and then we combine the
quality rankings into an overall ranking based on the relative priorities
between qualities.

\section{Results} \label{ch_results}

We initially identified 48 candidate software projects from the literature
\cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \cite{Emms2019,
Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  Then we filtered
as follows:

\item Removed the packages with no source code available, such as
\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.

\item Focused on MI software that provides visualization functions.  We removed
seven packages that were toolkits or libraries, such as \textit{VTK},
\textit{ITK}, and \textit{dcm4che}, and another three that were for PACS.

\item Removed \textit{Open Dicom Viewer} as it has not received any
updates since 2011.

@@

The Domain Expert provided a list 12 software packages.  We found 6 packages
were on both lists: \textit{3D Slicer}, \textit{Horos}, \textit{ImageJ},
\textit{Fiji}, \textit{MRIcron} (we use its descendant \textit{MRIcroGL}) and
\textit{Mango} (we use the web version \textit{Papaya}).  The remaining six
packages were on our out-of-scope list. The Domain Expert agreed with our final
choice of 29 packages -- see Table~\ref{tab_final_list}. This contains summary
data collected in the year 2020.

The projects are sorted in descending order of lines of code.  We found the
initial release dates (Rlsd) for most projects and marked the two unknown dates
with ``?''. The date of the last update is the date of the latest update, at the
time of measurement. We found funding information (Fnd) for only eight projects.
For the Number Of Contributors (NOC) we considered anyone who made at least one
accepted commit as a contributor. The NOC is not usually the same as the number
of long-term project members, since many projects received change requests and
code from the community.  With respect to the OS, 25 packages work on all three
OSs: Windows (W), macOS (M), and Linux (L). Although the usual approach to
cross-platform compatibility was to work natively on multiple OSes, five
projects achieved platform-independence via web applications. The full
measurement data for all packages is available on [removed for blind review]
%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
%% DOUBLE BLIND

\begin{table*}[!ht]
\centering
\begin{tabular}{p{3.7cm}lllllllll}
\toprule
\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cmidrule{7-9}
 &  &  &  &  &  & W & M & L &  \\ \midrule
ParaView \cite{Ahrens2005} & 2002 & 2020-10 & \checkmark & 100 & 886326 & \checkmark & \checkmark & \checkmark & \checkmark \\
Gwyddion \cite{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & \checkmark & \checkmark & \checkmark &  \\
Horos \cite{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & \checkmark &  &  \\
OsiriX Lite \cite{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & \checkmark &  &  \\
3D Slicer \cite{Kikinis2014} & 1998 & 2020-08 & \checkmark & 100 & 501451 & \checkmark & \checkmark & \checkmark &  \\
Drishti \cite{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & \checkmark & \checkmark & \checkmark &  \\
Ginkgo CADx \cite{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & \checkmark & \checkmark & \checkmark &  \\
GATE \cite{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & \checkmark & \checkmark &  \\
3DimViewer \cite{TESCAN2020} & ? & 2020-03 & \checkmark & 3 & 178065 & \checkmark & \checkmark &  &  \\
medInria \cite{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & \checkmark & \checkmark & \checkmark &  \\
BioImage Suite Web \cite{Papademetris2005} & 2018 & 2020-10 & \checkmark & 13 & 139699 &
\checkmark & \checkmark & \checkmark & \checkmark \\
Weasis \cite{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & \checkmark & \checkmark & \checkmark &  \\
AMIDE \cite{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & \checkmark & \checkmark & \checkmark &  \\
XMedCon \cite{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & \checkmark & \checkmark & \checkmark &  \\
ITK-SNAP \cite{Yushkevich2006} & 2006 & 2020-06 & \checkmark & 13 & 88530 & \checkmark & \checkmark & \checkmark &  \\
Papaya \cite{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & \checkmark & \checkmark & \checkmark &  \\
OHIF Viewer \cite{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & \checkmark & \checkmark & \checkmark & \checkmark \\
SMILI \cite{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & \checkmark & \checkmark & \checkmark &  \\
INVESALIUS 3 \cite{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & \checkmark & \checkmark & \checkmark &  \\
dwv \cite{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & \checkmark & \checkmark & \checkmark & \checkmark \\
DICOM Viewer \cite{Afsar2021} & 2018 & 2020-04 & \checkmark & 5 & 30761 & \checkmark & \checkmark & \checkmark &  \\
MicroView \cite{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & \checkmark & \checkmark & \checkmark &  \\
MatrixUser \cite{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & \checkmark & \checkmark & \checkmark &  \\
Slice:Drop \cite{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & \checkmark & \checkmark & \checkmark & \checkmark \\
dicompyler \cite{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & \checkmark & \checkmark &  &  \\
Fiji \cite{Schindelin2012} & 2011 & 2020-08 & \checkmark & 55 & 10833 & \checkmark & \checkmark & \checkmark &  \\
ImageJ \cite{Rueden2017} & 1997 & 2020-08 & \checkmark & 18 & 9681 & \checkmark & \checkmark & \checkmark &  \\
MRIcroGL \cite{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & \checkmark & \checkmark & \checkmark &  \\
DicomBrowser \cite{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & \checkmark & \checkmark & \checkmark &  \\ \bottomrule
\end{tabular}
\caption{Final software list (sorted in descending order of the number of Lines
Of Code (LOC))}
\label{tab_final_list}
\end{table*}

The programming languages used in order of decreasing popularity are \CC,
JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
for 11 of 29 projects; Pascal and Matlab were each used for a single project.

In the absence of
a specific real world context, we assumed all nine qualities are equally
important. Figure~\ref{fg_overall_scores} shows the overall scores in descending
order.

\begin{figure*}[!ht]
\includegraphics[scale=0.38]{overall_scores.pdf}
\caption{Overall AHP scores with an equal weighting for all 9 software qualities}

\label{fg_overall_scores}
\end{figure*}

[Summary description list]

We found
installation instructions for 16 projects, but two did not need them
(\textit{BioImage Suite Web} and \textit{Slice:Drop})
as they are web applications. 10 of the projects required extra
dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
\textit{DICOM Viewer} needs a Nextcloud platform.
